---
title: "Batch launches: hyperparameter sweeps and variant fleets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Batch launches: hyperparameter sweeps and variant fleets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
.vignette_db <- tempfile(fileext = ".duckdb")
options(modelrunnR.db = .vignette_db)
```

```{r setup}
library(modelrunnR)

# Toy data we'll sweep models over.
set.seed(1)
df <- data.frame(x = rnorm(50), y = rnorm(50))
stow(df, "train")
```

Most model work isn't one fit — it's a sweep. Three features times three
alphas, a few seeds, paired with a handful of preprocessing variants.
modelrunnR runs these as a single `launch()` call and returns one row
per run.

The toy "fits" below stand in for the kind of work you'd really sweep
(`glmnet`, `xgboost`, etc.) — the API is the same, just the body of
the `launch()` block changes.

## The helper

`mr_binds()` builds the sweep description. Each named argument is the
vector of values for that rebind slot.

```{r build-binds}
my_params <- mr_binds(
  alpha = c(0.1, 0.5, 1.0),
  beta  = c(0.0, 0.5, 1.0),
  mode  = "zip"
)
length(my_params)  # 3 envelopes (length-3 zipped with length-3)
```

`mode = "zip"` (default) pairs values element-wise. Lengths must line
up; length-1 slots recycle.

`mode = "cross"` takes the Cartesian product:

```{r cross-binds}
length(mr_binds(alpha = c(0.1, 1.0), beta = c(0, 0.5, 1), mode = "cross"))
# 6 envelopes (2 * 3)
```

## Running the batch

Hand the result to `rebind =`:

```{r run-batch}
result <- launch({
  d   <- grab("train") |> dplyr::collect()
  fit <- list(
    alpha = grab("alpha"),
    beta  = grab("beta"),
    coef  = with(d, cor(x, y)) * grab("alpha") + grab("beta")
  )
  stow(fit, "fit")
}, rebind = my_params)

nrow(result)
result[, c("status", "duration_ms", "variant_label")]
```

Each envelope writes its own `_mr_runs` row. The returned data frame
is the union, in envelope order.

## Sweeping over variant labels

`mr_variants()` is a small constructor for variant references — without
it, bare strings would be passed as literal rebind values rather than
variant refs:

```{r mr-variants, eval = FALSE}
mr_binds(features = mr_variants("clean", "raw", "imputed"),
         alpha    = c(0.1, 0.5, 1.0))
```

## Labels

Two ways to label batch runs: explicit per-envelope or
auto-propagation from upstream variants.

```{r explicit-labels}
labeled <- launch({
  fit <- list(alpha = grab("alpha"))
  stow(fit, "fit_labeled")
}, rebind = mr_binds(alpha = c(0.1, 0.5, 1.0),
                      .labels = c("low", "mid", "high")))
labeled$variant_label
```

For mixed per-envelope behavior (some explicit labels, some
auto-propagated), drop down to `mr_envelopes()`:

```{r envelopes, eval = FALSE}
mr_envelopes(
  list(.label = "baseline", features = mr_variant("clean"), alpha = 0.1),
  list(features = mr_variant("raw"), alpha = 1.0)            # auto-label
)
```

## Errors

A batch runs to completion. Failed envelopes write `status = "error"`
rows; the call raises (or warns, with `on_error = "warn"`) at the end.

```{r errors}
res <- tryCatch(
  launch({
    a <- grab("alpha")
    if (a == 0.5) stop("simulated failure")
    stow(list(a = a), "fit_err")
  }, rebind = mr_binds(alpha = c(0.1, 0.5, 1.0)),
  on_error = "warn"),
  warning = function(w) message("got: ", conditionMessage(w))
)

# 2 rows succeeded; 1 errored.
table(res$status)
```

## Seeing what each run bound to

Every batch run records its resolved rebinds on the run row as JSON.
The `value` field is human-readable: numbers stay as numbers, variant
names stay as names, data frames show `data.frame[<rows>x<cols>]`.

```{r rebinds}
recent <- DBI::dbGetQuery(mr_con(),
  "SELECT variant_label, rebinds FROM _mr_runs
    WHERE rebinds IS NOT NULL AND rebinds != '[]'
    ORDER BY started_at DESC
    LIMIT 3")

# Unpack the JSON for a more readable view.
do.call(rbind, lapply(seq_len(nrow(recent)), function(i) {
  rb <- jsonlite::fromJSON(recent$rebinds[i])
  data.frame(variant_label = recent$variant_label[i], rb)
}))
```

## SQL batches

The same dispatch works for SQL launches — useful when feature
engineering is one SQL body but the rebind varies across runs.

```{r sql-batch}
stow(data.frame(x = 1:5), "src")
stow(data.frame(x = 100:104), "src")    # second version

# `versions()` returns one row per stored version, latest first.
src_versions <- versions("src")
v_latest <- src_versions$content_hash[1]
v_older  <- src_versions$content_hash[2]

sql_batch <- launch(
  mr_sql("-- @inputs: src\n-- @output: src_summary\nSELECT COUNT(*) AS n FROM src"),
  rebind = mr_envelopes(
    list(.label = "older",  src = mr_hash(v_older)),
    list(.label = "latest", src = mr_hash(v_latest))
  )
)
sql_batch[, c("variant_label", "status")]
```

## Mental model

A batch launch is N independent single launches run back-to-back,
sharing nothing except the block of code and the `force` /
`duckdb_seed` / `on_error` flags. Staleness, label propagation, helper
tracking, and interactive-write warnings all behave per run exactly as
they do in single-launch mode.

```{r cleanup, include = FALSE}
modelrunnR:::.mr_reset_connection()
if (file.exists(.vignette_db)) file.remove(.vignette_db)
```
