---
title: "Getting started with modelrunnR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with modelrunnR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
# Isolate the DuckDB store to a tempdir so building this vignette
# doesn't leave a modelrunnR.duckdb file next to the package source.
.vignette_db <- tempfile(fileext = ".duckdb")
options(modelrunnR.db = .vignette_db)
```

```{r setup}
# install.packages("pak")
# pak::pak("jesseabrandt/modelrunnR")

library(modelrunnR)
```

## A tiny simulated dataset

We'll generate a toy regression problem — a linear signal plus noise —
so the rest of the vignette has something to model without shipping a
dataset.

```{r sim-data}
set.seed(42)
n  <- 200
df <- data.frame(x1 = rnorm(n), x2 = rnorm(n))
df$y <- 1 + df$x1 + 2 * df$x2 + rnorm(n, sd = 0.5)

head(df)
```

## Registering input data with `stow(mr_file(...))`

Before a tracked run can read input data, we need to register it with
the store. The verb is `stow()` — same as for any other value — and
file sources are tagged with `mr_file()` so `stow()` knows to route
them through the file-source path. The result is a content-addressed
versioned-shape version, reproducible across sessions because the
file's byte hash is tracked.

```{r register-input}
train_csv <- tempfile(fileext = ".csv")
write.csv(df, train_csv, row.names = FALSE)
stow(mr_file(train_csv), "training")
```

(`ingest("training", train_csv)` is the older spelling and still works
for one release cycle, with a deprecation warning.)

```{r grab-training}
# grab() returns a dbplyr lazy tbl for tabular data.
grab("training")

# collect() to materialize.
training_local <- grab("training") |> dplyr::collect()
head(training_local)
```

## Versioned-shape vs. append-shape vs. view-shape

The store has three storage shapes for tabular data, and which one you
get depends on how the value enters:

- **Versioned-shape — content-addressed versions.** `stow(mr_file(...))` writes
  input reference data as versioned-shape. Every distinct content gets its own
  version row; `grab()` returns the latest. Non-tabular values stowed
  via `stow(model, "fit")` also take this path (as artifacts).
- **Append-shape — run-indexed append log.** `stow(df, name)` *inside a
  tracked `launch()`* appends rows to a single growing DuckDB table,
  stamping each with the run's id and variant label. `grab(name)`
  returns one coherent snapshot by default: inside a launch it's the
  current run's rows; outside it's the latest run that wrote to the
  name, with system columns stripped. Pass `run = "all"` for the
  full cross-run view when you want to inspect history or compare
  runs.
- **View-shape — registered SQL view.** `stow(lazy_tbl, name, shape = "view")`
  wraps a `dbplyr` expression as a `CREATE OR REPLACE VIEW` rather than
  materializing it. Useful for rolling-window folds: pre-stow one view
  per fold (with `label =`), then rebind `train`/`test` per fold via
  `mr_variant()`. See `?stow` for a full example.

## A tracked run with `launch({ ... })`

Once you have something worth reproducing, wrap the modeling code in
`launch({ ... })`. The block runs inside an instrumented context:

- every `grab()` is recorded as an input,
- every `stow()` is recorded as an output,
- the expression's bytes, any declared external inputs, and
  wall-clock duration all land on a row in `_mr_runs`,
- a staleness line prints before the run so you know whether inputs
  or code have shifted since the last time you ran the same block.

**Give the pipeline a name.** As soon as you have something worth
iterating on, pass a `label`. A label is just a name for this
pipeline — *not* for this particular run. Every `launch()` with the
same label is a new *version* of that pipeline: same identity,
possibly different code. It's the unit modelrunnR uses to track
history, compare variants, and garbage-collect old versions later.

```{r launch}
run <- launch(
  {
    training <- grab("training") |> dplyr::collect()
    model    <- lm(y ~ x1 + x2, data = training)

    stow(model, "model")
    stow(data.frame(y = training$y, pred = predict(model, training)),
         "predictions")
  },
  label = "baseline"
)
```

`launch()` returns the run row invisibly.

```{r run-row}
run$status
run$run_id
run$variant_label
run$inputs
run$outputs
```

## Accumulating results across a sweep

A common pattern is sweeping a set of models and collecting their
metrics into a single, run-indexed table. `stow(df, name)` inside
`launch()` appends the data frame's rows to a append-shape accumulator
rather than creating a separate version per run — so every sweep run
contributes to one growing table you can compare all at once.

```{r sweep}
# A sweep across three models, all stowing "metrics" into the same
# Append-shape accumulator.
launch({
  fit <- lm(mpg ~ wt + hp, data = mtcars)
  tibble::tibble(
    model = "lm",
    rmse  = sqrt(mean(resid(fit)^2)),
    r2    = summary(fit)$r.squared
  ) |> stow("metrics")
}, label = "lm")

# Outside any launch, grab() returns the latest run's rows by default —
# one clean snapshot. To inspect the full sweep across runs, opt into
# the full-history view:
grab("metrics", run = "all") |> dplyr::collect()
```

## Inspecting versions

`versions()` walks the versioned-shape version history — one row per distinct
content hash. It works for versioned-input tables, SQL views, and stowed
artifacts. For append-shape names, rows live in the append table itself
(query it directly or use `grab()` outside `launch()`).

```{r versions}
versions("model")
```

Run the same block a second time and modelrunnR will *skip it*.
The expression bytes, the input hash, and the declared external
inputs all match the last run under this label, so there is
nothing to recompute. A row still lands on `_mr_runs` with
`status = "skipped_fresh"` so the provenance trail is intact, but
the block itself is not evaluated — pass `force = TRUE` (or set
`options(modelrunnR.skip_if_fresh = FALSE)`) to re-run anyway.

```{r rerun}
launch(
  {
    training <- grab("training") |> dplyr::collect()
    model    <- lm(y ~ x1 + x2, data = training)

    stow(model, "model")
    stow(data.frame(y = training$y, pred = predict(model, training)),
         "predictions")
  },
  label = "baseline"
)
```

## Iterating on the same pipeline

Now edit the modeling code — say, swap to a more flexible formula —
and re-run under the *same* label. modelrunnR records a new version
tied to that pipeline identity.

```{r iterate}
launch(
  {
    training <- grab("training") |> dplyr::collect()
    model    <- lm(y ~ x1 * x2, data = training)   # interaction added

    stow(model, "model")
    stow(data.frame(y = training$y, pred = predict(model, training)),
         "predictions")
  },
  label = "baseline"
)
```

`variants()` gives you the pipeline's history at a glance — one row
per labeled pipeline, with a run count and timestamps.

```{r variants}
variants(name = "predictions")
```

Three runs under `"baseline"`: the first launch, the skipped rerun,
and the edited iteration. Every `launch()` writes a `_mr_runs` row
(even skipped ones), and the row's `step` column keys on the
expression bytes — so the table shows two distinct inline steps
under the same label. When you're confident an older iteration
isn't worth keeping, `prune_variants()` and `prune()` can
clear the history (see `?prune_variants`, `?prune`).

## Inspecting runs

`runs()` returns the run log itself — the underlying `_mr_runs` table
as a tibble, one row per run. Use it when you want to drill into the
history beyond the pre-grouped views from `versions()` and
`variants()`.

```{r runs}
runs() |> dplyr::select(run_id, step, variant_label, status, started_at)
```

The `code_body` column carries an `mr_code` class, so pulling it out
prints as multi-line code (syntax-highlighted at a color-capable
terminal, plain text in pipes / files / knitr):

```{r runs-code}
runs() |>
  dplyr::filter(status == "success") |>
  utils::tail(1) |>
  dplyr::pull(code_body)
```

The class is purely R-side. `DBI::dbGetQuery()` against `_mr_runs`
returns plain `character`, so the store stays usable without the
package.

The remaining columns hold the rest of each run's context. The
JSON-shaped ones — `inputs`, `outputs`, `external_inputs`, `helpers`,
`rebinds`, `attached_packages` — are surfaced as plain `chr`; parse on
demand with `jsonlite::fromJSON()`.

`git_sha`, `git_branch`, and `git_dirty` stamp each row with the
working tree's commit at launch time so a run can be traced back to
the code that produced it. `git_dirty` is a short summary like
`"3 files changed, 47 insertions(+), 12 deletions(-)"` (plus an
untracked count when applicable) and is `NA` on a clean tree — its
presence is the dirty signal. All three are `NA` when the working
directory isn't a git repo or `git` isn't on `PATH`.

## Re-running a pipeline by label

Once a pipeline has a label, you don't have to re-type its code to
run it again. `launch(mr_label("baseline"))` looks up the most
recent run under that label and re-executes its code. For inline
pipelines the stored snapshot runs directly; for file pipelines the
file is re-sourced from disk (or the snapshot is used if the file
has been removed).

```{r relaunch}
launch(mr_label("baseline"))
```

The label auto-inherits, so the re-run stays in the same pipeline's
history.

## Staging runs without executing them

When you want to enumerate runs now and execute later — on an HPC submit
script, in a parallel `furrr` map, or just on tomorrow's run — `queue()`
stages the run row without firing the code. `queue()` registers a row
to `_mr_runs` with `status = "queued"`; `launch(mr_run(id))` picks it
up later and updates the row in place.

```{r queue-stage}
# Stage a single run against the same training data we used above.
q <- queue({
  training <- grab("training") |> dplyr::collect()
  model    <- lm(y ~ x1 + x2, data = training)
  stow(model, "queued_model")
})
q$status   # "queued"

# Pick it up — anywhere, any process. Here, in this session.
r <- launch(mr_run(q$run_id))
r$status   # "success" — same row, status flipped
identical(r$run_id, q$run_id)
```

Batch staging via `mr_binds()` writes N queued rows under one
`batch_id`:

```{r queue-batch, eval = FALSE}
qs <- queue(
  {
    training <- grab("training") |> dplyr::collect()
    model    <- lm(y ~ x1 + x2, data = training, weights = rep(grab("w"), nrow(training)))
    stow(model, "weighted_model")
  },
  rebind = mr_binds(w = c(0.5, 1.0, 2.0))
)

# Serial pickup:
purrr::walk(qs$run_id, ~ launch(mr_run(.x)))

# Or parallel via future / furrr — modelrunnR doesn't drive
# parallelism itself; compose it from the consumer side.
# furrr::future_map(qs$run_id, ~ launch(mr_run(.x)))
```

### Queueing a re-run

`queue()` accepts the same reference shapes as `launch()`'s relaunch
mode. Use `mr_label()` to stage a labeled pipeline's body for later
execution under different bindings:

```r
launch({ ... }, label = "ridge_pf")

queue(
  mr_label("ridge_pf"),
  rebind = mr_binds(panel = list(mr_variant("placebo_panel"))),
  label  = "ridge_pbo"
)
```

`mr_run(id)` works the same way against a specific historical run.

Pickup is in-place: `run_id`, `step`, `rebinds`, and `batch_id` stay
frozen across queue → pickup, while `status`, timing, and
session-context columns populate at execution. `code_body` is frozen
for inline steps; for file steps it's re-read from disk at pickup, with
a `warning()` if the file has drifted since queue time. If the freshness
check would skip the run, the queued row's `status` becomes
`"skipped_fresh"` without executing the body. See `?queue` for the
full contract.

## From block to script

`launch()` also accepts a path to a `.R` file. Once a block settles
into something you want in version control, move it to `fit.R` and
point `launch()` at the file instead — the label, inputs, outputs,
and staleness all work the same way.

```r
launch("fit.R", label = "baseline")
```

## SQL steps

When the work is feature engineering — joins, window functions, group
aggregates — DuckDB can do it natively. `launch()` accepts a `.sql`
file (or `mr_sql("...")` inline) and registers the result as a tracked
step alongside R-mode launches. The default is a **view** (cheap to
register; rows compute when something downstream reads them); pass
`materialize = TRUE` to force a `CREATE TABLE` instead.

The body is a bare `SELECT` (or `WITH ... SELECT`) — modelrunnR owns
the `CREATE OR REPLACE VIEW <physical> AS ...` wrapper. Optional
`-- @inputs:` and `-- @output:` headers declare which existing
modelrunnR names the SELECT references and what to call the result
(filename stem is the default for files; required for inline).

```{r sql-step}
sql_body <- "
  -- @inputs: training
  -- @output: training_summary
  SELECT
    AVG(y) AS mean_y,
    AVG(x1) AS mean_x1,
    AVG(x2) AS mean_x2,
    COUNT(*) AS n
  FROM training
"
launch(mr_sql(sql_body))

grab("training_summary") |> dplyr::collect()
```

Downstream R steps consume the view through the same `grab()` call as
any other tabular value:

```{r sql-then-r}
launch({
  s <- grab("training_summary") |> dplyr::collect()
  message(sprintf("training had n = %d rows, mean(y) = %.2f",
                  s$n, s$mean_y))
})
```

## Under the hood

A few things happen without you having to configure them.

**Where the DuckDB store lives.** modelrunnR walks up from your
working directory looking for a project marker — `DESCRIPTION`,
`*.Rproj`, `.git/`, `renv.lock`, or `.here` — and puts
`modelrunnR.duckdb` at that root. `setwd()` into a subdirectory and
you still hit the same DB, so a project has one store by default
rather than a scatter of little ones. If no marker is found, it
falls back to the current directory *and* warns you, so stray DBs
don't accumulate silently. `db_path()` shows you the resolved path,
and `options(modelrunnR.db = "...")` overrides it if you want to
point multiple projects at a shared store.

```{r cleanup, include = FALSE}
modelrunnR:::.mr_reset_connection()
if (file.exists(.vignette_db)) file.remove(.vignette_db)
```
