---
title: "Getting started with modelrunnR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with modelrunnR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
# Isolate the DuckDB store to a tempdir so building this vignette
# doesn't leave a modelrunnR.duckdb file next to the package source.
.vignette_db <- tempfile(fileext = ".duckdb")
options(modelrunnR.db = .vignette_db)
```

```{r setup}
# install.packages("pak")
# pak::pak("jesseabrandt/modelrunnR")

library(modelrunnR)
```

## A tiny simulated dataset

We'll generate a toy regression problem — a linear signal plus noise —
so the rest of the vignette has something to model without shipping a
dataset.

```{r sim-data}
set.seed(42)
n  <- 200
df <- data.frame(x1 = rnorm(n), x2 = rnorm(n))
df$y <- 1 + df$x1 + 2 * df$x2 + rnorm(n, sd = 0.5)

head(df)
```

## Stowing and grabbing at the REPL

`stow()` parks a value in the modelrunnR store under a logical name.
It's content-addressed: writing the same frame twice creates one
version row, not two. `grab()` is the other half — retrieve the latest
version (or a specific one) by name.

Even when called at the REPL, `stow()` records a row in `_mr_runs`
tagged as an interactive write so later tracked runs can flag inputs
that trace back to ad-hoc keyboard activity.

```{r stow-grab}
stow(df, "training")

# grab() returns a dbplyr lazy tbl for tabular data.
grab("training")

# collect() to materialize; DuckDB returns a tibble, so all.equal() checks values.
training_local <- grab("training") |> dplyr::collect()
all.equal(as.data.frame(training_local), df)
```

## A tracked run with `launch({ ... })`

Once you have something worth reproducing, wrap the modeling code in
`launch({ ... })`. The block runs inside an instrumented context:

- every `grab()` is recorded as an input,
- every `stow()` is recorded as an output,
- the expression's bytes, any declared external inputs, and
  wall-clock duration all land on a row in `_mr_runs`,
- a staleness line prints before the run so you know whether inputs
  or code have shifted since the last time you ran the same block.

**Give the pipeline a name.** As soon as you have something worth
iterating on, pass a `label`. A label is just a name for this
pipeline — *not* for this particular run. Every `launch()` with the
same label is a new *version* of that pipeline: same identity,
possibly different code. It's the unit modelrunnR uses to track
history, compare variants, and garbage-collect old versions later.

```{r launch}
run <- launch(
  {
    training <- grab("training") |> dplyr::collect()
    model    <- lm(y ~ x1 + x2, data = training)

    stow(model, "model")
    stow(data.frame(y = training$y, pred = predict(model, training)),
         "predictions")
  },
  label = "baseline"
)
```

`launch()` returns the run row invisibly.

```{r run-row}
run$status
run$run_id
run$variant_label
run$inputs
run$outputs
```

## Inspecting versions

```{r versions}
versions("predictions")
```

Run the same block a second time and modelrunnR will *skip it*.
The expression bytes, the input hash, and the declared external
inputs all match the last run under this label, so there is
nothing to recompute. A row still lands on `_mr_runs` with
`status = "skipped_fresh"` so the provenance trail is intact, but
the block itself is not evaluated — pass `force = TRUE` (or set
`options(modelrunnR.skip_if_fresh = FALSE)`) to re-run anyway.

```{r rerun}
launch(
  {
    training <- grab("training") |> dplyr::collect()
    model    <- lm(y ~ x1 + x2, data = training)

    stow(model, "model")
    stow(data.frame(y = training$y, pred = predict(model, training)),
         "predictions")
  },
  label = "baseline"
)
```

## Iterating on the same pipeline

Now edit the modeling code — say, swap to a more flexible formula —
and re-run under the *same* label. modelrunnR records a new version
tied to that pipeline identity.

```{r iterate}
launch(
  {
    training <- grab("training") |> dplyr::collect()
    model    <- lm(y ~ x1 * x2, data = training)   # interaction added

    stow(model, "model")
    stow(data.frame(y = training$y, pred = predict(model, training)),
         "predictions")
  },
  label = "baseline"
)
```

`variants()` gives you the pipeline's history at a glance — one row
per labeled pipeline, with a run count and timestamps.

```{r variants}
variants(name = "predictions")
```

Three runs under `"baseline"`: the first launch, the skipped rerun,
and the edited iteration. Every `launch()` writes a `_mr_runs` row
(even skipped ones), and the row's `step` column keys on the
expression bytes — so the table shows two distinct inline steps
under the same label. When you're confident an older iteration
isn't worth keeping, `prune_variants()` and `prune_versions()` can
clear the history (see `?prune_variants`, `?prune_versions`).

## Re-running a pipeline by label

Once a pipeline has a label, you don't have to re-type its code to
run it again. `launch(mr_label("baseline"))` looks up the most
recent run under that label and re-executes its code. For inline
pipelines the stored snapshot runs directly; for file pipelines the
file is re-sourced from disk (or the snapshot is used if the file
has been removed).

```{r relaunch}
launch(mr_label("baseline"))
```

The label auto-inherits, so the re-run stays in the same pipeline's
history.

## From block to script

`launch()` also accepts a path to a `.R` file. Once a block settles
into something you want in version control, move it to `fit.R` and
point `launch()` at the file instead — the label, inputs, outputs,
and staleness all work the same way.

```r
launch("fit.R", label = "baseline")
```

## SQL steps

When the work is feature engineering — joins, window functions, group
aggregates — DuckDB can do it natively. `launch()` accepts a `.sql`
file (or `mr_sql("...")` inline) and registers the result as a tracked
step alongside R-mode launches. The default is a **view** (cheap to
register; rows compute when something downstream reads them); pass
`materialize = TRUE` to force a `CREATE TABLE` instead.

The body is a bare `SELECT` (or `WITH ... SELECT`) — modelrunnR owns
the `CREATE OR REPLACE VIEW <physical> AS ...` wrapper. Optional
`-- @inputs:` and `-- @output:` headers declare which existing
modelrunnR names the SELECT references and what to call the result
(filename stem is the default for files; required for inline).

```{r sql-step}
sql_body <- "
  -- @inputs: training
  -- @output: training_summary
  SELECT
    AVG(y) AS mean_y,
    AVG(x1) AS mean_x1,
    AVG(x2) AS mean_x2,
    COUNT(*) AS n
  FROM training
"
launch(mr_sql(sql_body))

grab("training_summary") |> dplyr::collect()
```

Downstream R steps consume the view through the same `grab()` call as
any other tabular value:

```{r sql-then-r}
launch({
  s <- grab("training_summary") |> dplyr::collect()
  message(sprintf("training had n = %d rows, mean(y) = %.2f",
                  s$n, s$mean_y))
})
```

## Under the hood

A few things happen without you having to configure them.

**Where the DuckDB store lives.** modelrunnR walks up from your
working directory looking for a project marker — `DESCRIPTION`,
`*.Rproj`, `.git/`, `renv.lock`, or `.here` — and puts
`modelrunnR.duckdb` at that root. `setwd()` into a subdirectory and
you still hit the same DB, so a project has one store by default
rather than a scatter of little ones. If no marker is found, it
falls back to the current directory *and* warns you, so stray DBs
don't accumulate silently. `db_path()` shows you the resolved path,
and `options(modelrunnR.db = "...")` overrides it if you want to
point multiple projects at a shared store.

```{r cleanup, include = FALSE}
modelrunnR:::.mr_reset_connection()
if (file.exists(.vignette_db)) file.remove(.vignette_db)
```
