---
title: "Getting started with modelrunnR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with modelrunnR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
# Isolate the DuckDB store to a tempdir so building this vignette
# doesn't leave a modelrunnR.duckdb file next to the package source.
.vignette_db <- tempfile(fileext = ".duckdb")
options(modelrunnR.db = .vignette_db)
```

```{r setup}
# install.packages("pak")
# pak::pak("jesseabrandt/modelrunnR")

library(modelrunnR)
```

## A tiny simulated dataset

We'll generate a toy regression problem — a linear signal plus noise —
so the rest of the vignette has something to model without shipping a
dataset.

```{r sim-data}
set.seed(42)
n  <- 200
df <- data.frame(x1 = rnorm(n), x2 = rnorm(n))
df$y <- 1 + df$x1 + 2 * df$x2 + rnorm(n, sd = 0.5)

head(df)
```

## Stowing and grabbing at the REPL

`stow()` parks a value in the modelrunnR store under a logical name.
It's content-addressed: writing the same frame twice creates one
version row, not two. `grab()` is the other half — retrieve the latest
version (or a specific one) by name.

Even when called at the REPL, `stow()` records a row in `_mr_runs`
tagged as an interactive write so later tracked runs can flag inputs
that trace back to ad-hoc keyboard activity.

```{r stow-grab}
stow(df, "training")

training <- grab("training")
identical(training, df)
```

## A tracked run with `launch({ ... })`

Once you have something worth reproducing, wrap the modeling code in
`launch({ ... })`. The block runs inside an instrumented context:

- every `grab()` is recorded as an input,
- every `stow()` is recorded as an output,
- the expression's bytes, any declared external inputs, and
  wall-clock duration all land on a row in `_mr_runs`,
- a staleness line prints before the run so you know whether inputs
  or code have shifted since the last time you ran the same block.

```{r launch}
run <- launch({
  training <- grab("training")
  model    <- lm(y ~ x1 + x2, data = training)

  stow(model, "model")
  stow(data.frame(y = training$y, pred = predict(model, training)),
       "predictions")
})
```

`launch()` returns the run row invisibly.

```{r run-row}
run$status
run$run_id
run$inputs
run$outputs
```

## Inspecting versions

```{r versions}
versions("predictions")
```

Run the same block a second time and modelrunnR will report it as
fresh — the expression bytes and the input hash haven't moved, so no
new version is written.

```{r rerun}
launch({
  training <- grab("training")
  model    <- lm(y ~ x1 + x2, data = training)

  stow(model, "model")
  stow(data.frame(y = training$y, pred = predict(model, training)),
       "predictions")
})
```

## From block to script

`launch()` also accepts a path to a `.R` file. Once a block settles
into something you want in version control, move it to `fit.R` and
point `launch()` at the file instead — everything else (inputs,
outputs, staleness, rebind, labels) works the same way.

```r
launch("fit.R")
```

```{r cleanup, include = FALSE}
modelrunnR:::.mr_reset_connection()
if (file.exists(.vignette_db)) file.remove(.vignette_db)
```
