---
title: "Lazy data with modelrunnR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Lazy data with modelrunnR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
.vignette_db <- tempfile(fileext = ".duckdb")
options(modelrunnR.db = .vignette_db)
```

```{r setup}
library(modelrunnR)
library(dplyr)
```

## `grab()` returns a lazy reference

modelrunnR stores tables in DuckDB. When you `grab()` a tabular value, you
get a `dbplyr` lazy reference back — nothing is pulled into R memory yet.
This is what makes the `read_csv()`-replacement story hold up for files
that don't fit in RAM.

```{r sim}
# Simulate a moderately-sized dataset on disk as a CSV.
tmp <- tempfile(fileext = ".csv")
write.csv(
  data.frame(
    region = sample(c("N", "S", "E", "W"), 50000, replace = TRUE),
    year   = sample(2018:2024, 50000, replace = TRUE),
    price  = rlnorm(50000, meanlog = 11, sdlog = 0.5),
    sqft   = round(rnorm(50000, mean = 1800, sd = 400))
  ),
  tmp, row.names = FALSE
)

raw <- grab("sales", source = tmp)
class(raw)
```

`grab()` ingested the CSV server-side via DuckDB's `read_csv_auto()` —
no R-side materialization. Subsequent `grab("sales")` calls return a
lazy tbl directly.

## Composing `dplyr` verbs stays lazy

```{r verbs}
raw |>
  filter(year >= 2020) |>
  group_by(region) |>
  summarise(n = n(), avg_price = mean(price), .groups = "drop")
```

That ran as SQL inside DuckDB. Use `show_query()` to see exactly what
DuckDB executed:

```{r show-query}
raw |>
  filter(year >= 2020) |>
  group_by(region) |>
  summarise(avg_price = mean(price), .groups = "drop") |>
  show_query()
```

## Materializing

Three idioms — pick whichever reads most naturally:

```{r collect}
# dplyr::collect()
sample1 <- raw |> slice_sample(n = 100) |> collect()

# as.data.frame()
sample2 <- as.data.frame(raw |> slice_sample(n = 100))

# tibble::as_tibble()
sample3 <- tibble::as_tibble(raw |> slice_sample(n = 100))
```

Non-`dplyr` consumers that coerce via `as.data.frame()` (most of base
`stats`, `ggplot2`) auto-collect transparently:

```{r stats}
fit <- lm(price ~ sqft, data = raw |> slice_sample(n = 1000) |> collect())
coef(fit)
```

Two things *don't* auto-collect: `x$col` (base `$` isn't defined on
`tbl_lazy`) and base `[` subsetting. Use `dplyr::pull(col)` or collect
first.

## Reproducible sampling with `duckdb_seed`

`dplyr::slice_sample()` against a lazy tbl uses DuckDB's RNG — R's
`set.seed()` does not reach it. Pass `duckdb_seed` to `launch()`:

```{r seed-run}
launch(
  {
    sample <- grab("sales") |>
      slice_sample(n = 1000) |>
      collect()
    stow(sample, "sales_sample")
  },
  label   = "sample_demo",
  duckdb_seed = 0.42
)
```

Every `launch()` under the same `duckdb_seed` produces a byte-identical
sample. Valid range is `[-1, 1]`; the value is recorded on the run
row for reproducibility audits.

## Whole-step server-side pipelines

`stow()` accepts a lazy tbl. The entire pipeline step runs as SQL — no
R-side materialization, even at the end:

```{r lazy-stow}
launch(
  {
    grab("sales") |>
      filter(year == 2024) |>
      group_by(region) |>
      summarise(total = sum(price), .groups = "drop") |>
      stow("sales_by_region_2024")
  },
  label = "regional_rollup"
)

# Downstream: grab() returns another lazy tbl — composes right back in.
grab("sales_by_region_2024") |> collect()
```

The rendered SQL is stored alongside the output on `_mr_versions.source_sql`
for later inspection.

## Escape hatches

For sampling schemes `dbplyr` doesn't express — stratified, grouped CV,
custom — two reliable exits:

**Raw SQL via `mr_con()`.** The logical name (`sales`) is a view over
the latest version, so you can query it directly:

```{r raw-sql}
DBI::dbGetQuery(
  mr_con(),
  "SELECT region, COUNT(*) AS n
   FROM (SELECT * FROM sales USING SAMPLE 5 PERCENT (reservoir, 42)) s
   GROUP BY region"
)
```

That's DuckDB's reservoir sample (seeded inline) aggregated by region —
one way to express sampling schemes `dbplyr` doesn't cover.

**Collect then sample in R.** When the sampling logic is cleaner in R
— grouped CV, rolling windows, stratification — pull the keys you
need, sample in R, stow the partition:

```{r r-side-index}
# Pull just the region + year columns (cheap); build folds in R.
keys <- grab("sales") |>
  select(region, year) |>
  collect() |>
  mutate(rid = row_number())
# ... R-side sampling / fold construction over `keys` ...
head(keys)
```

Use whichever fits the structure you're building.
