---
title: "Nested sweeps: hyperparameter tuning and k-fold CV"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nested sweeps: hyperparameter tuning and k-fold CV}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
.vignette_db <- tempfile(fileext = ".duckdb")
options(modelrunnR.db = .vignette_db)
```

Cross-validating three hyperparameter values across five folds is
fifteen model fits arranged in a 3 × 5 grid. Conceptually that's a
nested loop; mechanically, modelrunnR runs it as a flat cross-product
of rebind envelopes and lets the append-shape accumulator plus a
post-batch aggregation launch do the "nesting" for you — no nested
`launch()` calls.

See `vignette("batch-launches")` for the general feature; this is a
cookbook application of it.

## The shape

- **Outer axis — hyperparameter.** One candidate value per rebind, e.g.
  `alpha ∈ {0.1, 0.5, 1.0}`.
- **Inner axis — fold.** One value per held-out partition, e.g.
  `fold ∈ 1:5`.
- **Flat sweep.** `mode = "cross"` on `mr_binds()` expands the two axes
  into a single batch of 15 envelopes. Each envelope is an independent
  `launch()` run writing a one-row metrics result to a shared
  append-shape table.
- **Aggregation.** A second `launch()` reads the 15-row accumulator,
  groups by `alpha`, and writes a 3-row CV summary.

One `batch_id` ties the 15 runs together, one append-shape table holds
their metrics, and one follow-up run computes mean-RMSE per `alpha`.

## Setup

A toy regression problem so the rest of the vignette has something to
fit:

```{r setup}
library(modelrunnR)

set.seed(42)
n  <- 200
df <- data.frame(x1 = rnorm(n), x2 = rnorm(n))
df$y <- 1 + df$x1 + 2 * df$x2 + rnorm(n, sd = 0.5)

# Register as an ingested source so launch() blocks can grab() it.
train_csv <- tempfile(fileext = ".csv")
write.csv(df, train_csv, row.names = FALSE)
ingest("training", train_csv)
```

## Pre-assign fold IDs

If every envelope sampled its own random split, per-alpha RMSEs across
folds wouldn't be directly comparable. Assign fold IDs once, stow them,
and let every sweep run grab the same splits:

```{r folds}
launch({
  dat <- grab("training") |> dplyr::collect()
  set.seed(1)
  dat$fold_id <- sample(1:5, nrow(dat), replace = TRUE)
  stow(dat, "folds")
}, label = "make-folds")
```

## The nested sweep

One call. 3 × 5 = 15 envelopes, one run row per envelope, all tied
together by a shared `batch_id`.

`mr_binds(mode = "cross")` expands with the leftmost argument varying
fastest, matching `expand.grid()` — so computing per-envelope labels is
a one-liner:

```{r sweep}
grid <- expand.grid(alpha = c(0.1, 0.5, 1.0), fold = 1:5)

params <- mr_binds(
  alpha   = c(0.1, 0.5, 1.0),
  fold    = 1:5,
  mode    = "cross",
  .labels = sprintf("alpha=%s_fold=%d", grid$alpha, grid$fold)
)

launch({
  # Inside a launch, append-shape grabs default to the *current* run's
  # rows — designed for the write-then-read-back pattern. Here we're
  # reading data a *prior* run stowed, so ask for it by source variant:
  folds <- grab("folds", variant = "make-folds") |> dplyr::collect()
  train <- dplyr::filter(folds, fold_id != grab("fold"))
  held  <- dplyr::filter(folds, fold_id == grab("fold"))

  # Toy "model": fit unshrunk, then shrink coefficients toward zero by
  # alpha. Stand-in for a real regularized fit (glmnet, xgboost, etc.).
  fit      <- lm(y ~ x1 + x2, data = train)
  shrunken <- coef(fit) * (1 - grab("alpha"))
  pred     <- cbind(1, held$x1, held$x2) %*% shrunken

  tibble::tibble(
    alpha = grab("alpha"),
    fold  = grab("fold"),
    rmse  = sqrt(mean((pred - held$y)^2))
  ) |> stow("cv_metrics")
}, rebind = params)
```

Four things to notice:

- **`.labels` gives every envelope its own variant identity.** Without
  them, every envelope runs the same code against the same grabs, and
  the skip-on-fresh check — which is keyed on `(step, label)` —
  considers the second envelope a repeat of the first. Unique labels
  per sweep coordinate keep each envelope independently tracked.
- **`grab("folds", variant = "make-folds")`.** Inside a launch, the
  append-shape default returns only the current run's rows. The
  `variant =` selector reads the latest rows stowed under that source
  variant — i.e. the fold assignments from the prep step.
- **`alpha` and `fold` are carried in the stowed tibble.** Their values
  are on the run row via `_mr_runs.rebinds` already, but surfacing them
  as columns on the append-shape `cv_metrics` table makes the
  aggregation step a plain `group_by()` — no JSON parsing. See
  `TODO.md` for a follow-up that would auto-surface rebind values.
- **No nested `launch()`.** The sweep block runs straight through to
  `stow()`. `fold` is just another dimension of the cross-product, not
  an inner loop.

The batch returns a 15-row data frame shaped like the single-run return:

```{r batch-return, eval = FALSE}
# Each row is one envelope; columns include run_id, status,
# duration_ms, variant_label, and the rebinds JSON.
```

And the accumulator can be inspected directly:

```{r inspect-accumulator}
grab("cv_metrics", run = "all") |> dplyr::collect()
```

## Aggregating fold-level metrics

A second `launch()` reads the 15-row accumulator back as a single lazy
tbl and summarises it. `run = "all"` is the opt-in to the full-history
view; without it, a `grab()` outside the originating launch would
return only the latest run's rows.

```{r aggregate}
launch({
  grab("cv_metrics", run = "all") |>
    dplyr::group_by(alpha) |>
    dplyr::summarise(
      mean_rmse = mean(rmse),
      sd_rmse   = sd(rmse),
      .groups   = "drop"
    ) |>
    stow("cv_summary")
}, label = "cv-summary")

grab("cv_summary") |> dplyr::collect()
```

`cv_summary` is itself append-shape, so re-running the sweep (e.g. with
`alpha = c(0.1, 0.5, 1.0, 1.5)`) and re-running this aggregation writes
a fresh row set keyed by the new `run_id`.

## Recovering a batch

Every envelope in a batch shares a `batch_id`. That's the anchor when
you want to query the sweep as a group — not just by label, but as
"the N runs from that call":

```{r batch-id}
DBI::dbGetQuery(mr_con(),
  "SELECT batch_id, COUNT(*) AS n_envelopes, MIN(started_at) AS started
     FROM _mr_runs
    WHERE batch_id IS NOT NULL
    GROUP BY batch_id ORDER BY started DESC")
```

## Mental model

- Nested loops are a cognitive frame. The package runs them flat.
- `mr_binds(..., mode = "cross")` turns the nesting into one sequence
  of envelopes.
- The append-shape accumulator captures per-envelope results in one
  growing table.
- A second `launch()` aggregates over that table. This is the "inner
  reduction" of the logical nested loop — but mechanically, it's just
  another tracked run.

Three tracked runs hold the full workflow: `make-folds`, the 15-run
batch sweep, and `cv-summary`. Everything else is flat.

```{r cleanup, include = FALSE}
modelrunnR:::.mr_reset_connection()
if (file.exists(.vignette_db)) file.remove(.vignette_db)
```
